---
title: 'Project 1: Airlines'
author: 'Michał Lubczyński'
date: '01.04.2022'
output:
  pdf_document: default
---


```{r include = FALSE}
library(DBI)
library(kableExtra)
db <- 'dwbd_flights' 
host_db <- # University database
db_port <- 5432
db_user <- rstudioapi::askForPassword("Database user")
db_password <- rstudioapi::askForPassword("Database password")
con <- dbConnect(RPostgres::Postgres(), dbname = db, host=host_db, port=db_port, user=db_user, password=db_password)
```


**a) What was the average arrival delay?**\
```{sql connection=con, output.var = "a", echo = TRUE, results = "hide"}
SELECT avg(arr_delay_new) as avg_delay
FROM "Flight_delays"
```
```{r echo = FALSE}
kableExtra::kable(a, booktabs = T) %>%
  row_spec(0, bold = T) %>%
  kable_styling(latex_options = c("hold_position"))
```


**b) What was the maximum arrival delay?**\
```{sql connection=con, output.var = "b", echo = TRUE, results = "hide"}
SELECT max(arr_delay_new)
FROM "Flight_delays"
```
```{r echo = FALSE}
kableExtra::kable(b, booktabs = T) %>%
  row_spec(0, bold = T) %>%
  kable_styling(latex_options = c("hold_position"))
```


**c) Which flight had the biggest delay in arrival?**\
```{sql connection=con, output.var = "c", echo = TRUE, results = "show"}
SELECT carrier, 
        origin_city_name,
        dest_city_name,
        fl_date,
        arr_delay_new
FROM "Flight_delays"
WHERE arr_delay_new IN (SELECT MAX(arr_delay_new)
                            FROM "Flight_delays")
```
```{r echo = FALSE}
kableExtra::kable(c, booktabs = T) %>%
  row_spec(0, bold = T) %>%
  kable_styling(latex_options = c("hold_position")) 
```


**d) Which days of the week are the worst to travel?**\
```{sql connection=con, output.var = "d", echo = TRUE, results = "hide"}
SELECT weekday_name, 
        AVG(arr_delay_new) AS avg_delay
FROM "Weekdays" 
   JOIN "Flight_delays" ON weekday_id = day_of_week
GROUP BY weekday_name 
ORDER BY avg_delay DESC
```
```{r echo = FALSE}
kableExtra::kable(d, booktabs = T) %>%
  row_spec(0, bold = T) %>%
  kable_styling(latex_options = c("hold_position")) 
```


**e) Which airlines flying from San Francisco (SFO) have the lowest arrival delay?**\
```{sql connection=con, output.var = "e", echo = TRUE, results = "hide"}
SELECT AL.airline_name, 
        avg(FD.arr_delay_new) AS avg_delay
FROM "Flight_delays" FD
    JOIN "Airlines" AL ON FD.airline_id = AL.airline_id 
WHERE FD.airline_id IN (SELECT airline_id 
                        FROM "Flight_delays"
                        WHERE origin = 'SFO')
GROUP BY airline_name
ORDER BY avg_delay DESC
```
```{r echo = FALSE}
kableExtra::kable(e, booktabs = T) %>%
  row_spec(0, bold = T) %>%
  kable_styling(latex_options = c("hold_position")) 
```


**f) What proportion of airlines has regular delays, i.e., their flight has an average delay of at least 10 minutes?**\
```{sql connection=con, output.var = "f", echo = TRUE, results = "hide"}
SELECT (count(DISTINCT FD2.airline_id)*1.0)
        /count(DISTINCT FD1.airline_id) AS late_proportion
FROM "Flight_delays" FD1
    JOIN "Flight_delays" FD2 ON FD1.fl_num = FD2.fl_num
    WHERE FD2.airline_id IN (SELECT airline_id
                                FROM "Flight_delays"
                                GROUP BY airline_id
                                HAVING AVG(arr_delay_new)>10)
```
```{r echo = FALSE}
kableExtra::kable(f, booktabs = T) %>%
  row_spec(0, bold = T) %>%
  kable_styling(latex_options = c("hold_position")) 
```


**g) How do departure delays affect arrival delays?**\
Pearson correlation coefficient between departure delay time and arrival delay time.\
```{sql connection=con, output.var = "g", echo = TRUE, results = "hide"}
SELECT ((SUM(dep_delay_new * arr_delay_new)
     - (SUM(dep_delay_new) * SUM(arr_delay_new)) / COUNT(*)
    )
   )
   / (SQRT(SUM(dep_delay_new * dep_delay_new)
           - (SUM(dep_delay_new) * SUM (dep_delay_new)) / COUNT(*)
          )
      * SQRT(SUM(arr_delay_new * arr_delay_new)
             - (SUM(arr_delay_new) * SUM(arr_delay_new)) / COUNT(*)
            )
     ) AS Pearsons_r
FROM "Flight_delays"
```
```{r echo = FALSE}
kableExtra::kable(g, booktabs = T) %>%
  row_spec(0, bold = T) %>%
  kable_styling(latex_options = c("hold_position")) 
```


**h) Which airline had the largest increase in average arrival delay during the last week of the month, i.e., between July 1-23 and July 24-31?**\
```{sql connection=con, output.var = "h", echo = TRUE, results = "hide"}
SELECT AL.airline_name, 
        avg(FD1.arr_delay_new)
        -avg(FD2.arr_delay_new) AS delay_increase
FROM "Flight_delays" FD1
    JOIN "Flight_delays" FD2 ON FD1.airline_id = FD2.airline_id
    AND FD2.day_of_month<24 AND FD1.day_of_month>=24
    JOIN "Airlines" AL ON FD1.airline_id = AL.airline_id 
GROUP BY AL.airline_name
ORDER BY delay_increase DESC 
LIMIT 1
```
```{r echo = FALSE}
kableExtra::kable(h, booktabs = T) %>%
  row_spec(0, bold = T) %>%
  kable_styling(latex_options = c("hold_position")) 
```


**i) Which airlines fly both SFO -> PDX (Portland) and SFO -> EUG (Eugene)?**\
```{sql connection=con, output.var = "i", echo = TRUE, results = "hide"}
SELECT DISTINCT airline_name
FROM "Airlines" AL
    Left JOIN "Flight_delays" FD ON AL.airline_id = FD.airline_id
    AND FD.dest ='EUG'
WHERE FD.origin = 'SFO' AND AL.airline_id IN (SELECT airline_id
                                                FROM "Flight_delays"
                                                WHERE dest = 'PDX')
```
```{r echo = FALSE}
kableExtra::kable(i, booktabs = T) %>%
  row_spec(0, bold = T) %>%
  kable_styling(latex_options = c("hold_position")) 
```


**j) What is the fastest way to get from Chicago to Stanford, assuming departure after 2pm local time?**\
```{sql connection=con, output.var = "j", echo = TRUE, results = "hide"}
SELECT origin,
        dest,
        avg(arr_delay_new) AS avg_delay
FROM "Flight_delays" 
WHERE origin IN ('MDW','ORD') 
AND dest IN ('SFO','SJC','OAK')
AND crs_dep_time > 1400
GROUP BY origin,
        dest
ORDER BY avg_delay desc
```
```{r echo = FALSE}
kableExtra::kable(j, booktabs = T) %>%
  row_spec(0, bold = T) %>%
  kable_styling(latex_options = c("hold_position")) 
```